Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-11-05 Manuscript/Text/manuscript.tex
Add GEO submission number

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms. Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}. Numerous studies on the transcriptomic effects of virus infection in model organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}. Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways, such as RNA silencing, Toll pathways, IMD pathways, JAK/STAT pathways, autophagy, and endocytosis \cite{McMenamin}.

@@

We next used repLIcate TREatment (``litre'') plots, which we recently developed and published in our bigPint software package. Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix. In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups. For example, there are nine ways to pair a replicate from one treatment group with a replicate from the other treatment group in the Galbraith dataset (N.1 and V.1, N.1 and V.2, N.1 and V.3, N.2 and V.1, N.2. and V.2, N.2 and V.3, N.3 and V.1, N.3 and V.2, and N.3 and V.3). Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot. With 11,825 genes in the Galbriath data, 106,425 points would need to be plotted. Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points. For either dataset, plotting all these points would reduce the speed of interactive functionality and cause overplotting problems. As a result, we use hexagon bins to summarize this massive information. Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.

Additional file 4 shows nine example litre plots for our dataset. The hexagon background is the same for all nine litre plots because it simply shows the distribution of all between-treatment sample pair combinations for \textit{all} genes in our dataset. In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest. Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith datset. We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbriath data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3). Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset. We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6). We also observe that, in the Galbraith dataset, the DEG points in the first cluster show less tight cluster patterns than the DEG points in the second cluster (Additional files 5 and 6), an observation we saw previously in the parallel coordinate plots (Figure 3).

@@

One of the few studies that has investigated transcriptomic response to nutrition in honey bees similarly found that pollen upregulates genes related to macromolecule metabolism, insulin pathways, and TOR pathways \cite{alaux2}. Diet effects on transcriptomics have been more extensively studied in the insect model Drosophila. One recent transcriptomic study in \textit{Drosophila melanogaster} reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. \textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. As we saw in our study, these studies generally suggest that diet differences may relate to gene expression changes in metabolism and immune responses in honey bees.

@@

There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees. These studies have shown that RNA silencing is a major antiviral strategy, along with transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7-autophagy pathways \cite{cherryReview, swevers}. Recent transcriptomic studies in honey bees have shown similar hallmarks of these same antiviral defense mechanisms \cite{McMenamin}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived from the diverse methods of replication and infection cycle evolved by different viruses. An intricate set of pro- and anti-virus host factors such as ribosomal proteins and autophagy pathways are involved, but the response depends on the virus species, as has been elucidated in Drosophila \cite{cherryReview, swevers}. In addition, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken, Brutscher}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

\section*{Availability of data and materials}
  The datasets generated and/or analysed during the current study are available in the [NAME] repository, [PERSISTENT WEB LINK TO DATASETS]. Include our data, Galbraith data, scripts to reproduce tables and figures (on GitHub).

@@

\section*{Author's contributions}
  LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript. BB conceptualized the study and critically revised the manuscript. AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq. 
DC advised on statistical analyses and visualisation.
    %All authors read and approved the final manuscript.

>>>>

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms. Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}. Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}. Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways \cite{McMenamin}.

@@

We next used repLIcate TREatment (``litre'') plots, which we recently developed and published in our bigPint software package. Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix. In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups. For example, there are nine ways to pair a replicate from one treatment group with a replicate from the other treatment group in the Galbraith dataset (N.1 and V.1, N.1 and V.2, N.1 and V.3, N.2 and V.1, N.2. and V.2, N.2 and V.3, N.3 and V.1, N.3 and V.2, and N.3 and V.3). Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot. With 11,825 genes in the Galbraith data, 106,425 points would need to be plotted. Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points. For either dataset, plotting all these points would reduce the speed of interactive functionality and cause overplotting problems. As a result, we use hexagon bins to summarize this massive information. Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.

Additional file 4 shows nine example litre plots for our dataset. The hexagon background is the same for all nine litre plots because it simply shows the distribution of all between-treatment sample pair combinations for \textit{all} genes in our dataset. In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest. Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith dataset. We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbraith data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3). Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset. We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6). We also observe that, in the Galbraith dataset, the DEG points in the first cluster show less tight cluster patterns than the DEG points in the second cluster (Additional files 5 and 6), an observation we saw previously in the parallel coordinate plots (Figure 3).

@@

One of the few studies that has investigated transcriptomic response to nutrition in honey bees similarly found that pollen upregulates genes related to macromolecule metabolism, insulin pathways, and TOR pathways \cite{alaux2}. Diet effects on transcriptomics have been more extensively studied in the insect model Drosophila. One recent transcriptomic study in \textit{Drosophila melanogaster} reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. \textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in response to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. As we saw in our study, these studies generally suggest that diet differences may relate to gene expression changes in metabolism and immune responses in honey bees.

@@

There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees. These studies have showed that RNA silencing is a major antiviral strategy, along with transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7-autophagy pathways \cite{cherryReview, swevers}. Recent transcriptomic studies in honey bees have shown similar hallmarks of these same antiviral defense mechanisms, including RNA silencing, Toll pathways, IMD pathways, JAK/STAT pathways, autophagy, and endocytosis \cite{McMenamin}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived from the diverse methods of replication and infection cycle evolved by different viruses. An intricate set of pro- and anti-virus host factors such as ribosomal proteins and autophagy pathways are involved, but the response depends on the virus species, as has been elucidated in Drosophila \cite{cherryReview, swevers}. In addition, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken, Brutscher}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

\section*{Availability of data and materials}
  The data discussed in this publication have been deposited in NCBI's Gene Expression Omnibus \cite{geo} and are accessible through GEO Series accession number GSE121885 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE121885). The scripts to reproduce analyses and figures in this publication are available online (https://github.com/lrutter/HoneyBeePaper).

@@

\section*{Author's contributions}
  LR performed the bioinformatic and statistical analyses, produced the figures and tables, and drafted the manuscript. BB conceptualized the study and critically revised the manuscript. AD contributed to experimental design, carried out the laboratory experiments, and processed samples for virus titers and RNA-seq. 
DC advised on statistical analyses and visualization.
    %All authors read and approved the final manuscript.
